---
title: "Customize Block Shape of Chunk Code in R Markdown"
author: "M Adi Setiawan"
date: "2023-09-05"
output: html_document
---


## GET STARTED WITH CUSTOM BLOCKS CODE WITH CSS STYLE

```{r standar chunk shape format, include=TRUE}

print ('this is a standar output of chunk block')

```

if you want to formatting chunk block, you should use css.style, below steps how to use css.style :

1. first you need, decleare custom CSS rules : 

``` {css chunk style, echo=FALSE}
.watch-out {
  background-color: #24293E;
  border: 1px solid #9336B4 ;
  color : white;
}
```

2. Then we assign a class `watch-out` to the code chunk via the chunk option `class.source`.

```{r class.source = "watch-out", class.output="watch-out" , include=TRUE}

print ('this is a format chunk block')

```

3. If you want all code blocks in the document to use the custom style, you can set class.source in the global knitr options, e.g.,

```{r, echo=TRUE, include=TRUE, results='hide'}
knitr::opts_chunk$set(class.source = "watch-out") # applying all custom style on input/source of chunk code
knitr::opts_chunk$set(class.output = "watch-out") # applying all custom style on output of chunk code
```

```{r, include = TRUE}
print ('result of applying css style to input & output block of chunk code')
```



### B. LIST PROPERTIES FOR CUSTOMIZING CODE BLOCKS WITH CSS STYLE

here's list you can use : 

* background-color: using for filling background color
* border: using for format border line : such as : line width, line color
* color : adjusting color text
* text-align : (left, center, right)
* font-family: 'Monaco', monospace; (example)
* font-size: adjusting font size (combined with px)
* line-height: adjusting for text (combined with %)
* padding : adjusting border height (combined with ex)
* border-radius: formating border shape (combined px)

 ```
 will explain later :
  * letter-spacing: normal
  * word-break: break-all
  ```

``` {css chunk style-2, echo=FALSE}
.variety {
  background-color: #071952;
  border: 2px solid #8BE8E5 ;
  color : white;
  text-align : center ;
  font-family: 'Monaco', monospace;
  font-size: 12px ;
  letter-spacing: normal;
  word-break: break-all;
  letter-spacing: -0.05em;
  word-break: normal;
  border-radius: 15px;
  line-height: 100%;
  padding: 1ex;
}

```

```{r class.source = "variety", class.output="variety" , include=TRUE}

print ('this code block using variety format')

```






we will use basic css style below :
``` {css chunck style-3, echo=FALSE}
.variety {
  background-color: #57375D;
  border: 1px solid #9400FF ;
  color : white;
}
```

```{r basic,class.source = "variety", class.output="variety" ,include=TRUE}

print ('this is a basic code block')

```