---
title: "Plot Layer of ggplot"
author: "M Adi Setiawan"
date: "2023-09-10"
output: html_document
---
reff :
  1. Theory : https://ggplot2-book.org/introduction
  2. Aes Spesifications : https://ggplot2.tidyverse.org/articles/ggplot2-specs.html

```{r library, include=FALSE}
library(ggplot2)
```

```{r raw data, include=FALSE}
raw_data <- mtcars
```

> tilde (~) : to separate the left hand side of an equation from the right hand side.
* left side of the tilde operator (y) represents the response variable. (dependent var)
* right side of the tilde operator (x1, x2) represent the predictor variables. (independent var)
* complete graph should have : data, mapping aes, labs (x=,y=) , theme (title), xlim(x1,x2) and ylim (y1,y2) 

# 1 Getting Started

## A. Basic Layer of ggplot

ggplot(data = NULL, mapping = aes(), ..., environment = parent.frame()

* Key components of ggplot : 
  
  1. data :  A dataset which overrides the default plot dataset.
  
  2. aes / aesthetics : A set of aesthetic mappings between variables in the data and visual properties
  
  3. additional layer : At least one layer which describes how to render each observation. Layers are usually created with a geom function.

* for more detail documentation about ggplot :  library(help = "ggplot2" )

```{r basic ggplot, include=FALSE}
library (ggplot2)
data <- mtcars

# basic ggplot

ggplot(data = data, mapping = aes(x = mpg, y = disp) ) + geom_point()

#This produces a scatterplot defined by:

# Data: data = mtcars
# Aesthetic mapping: mpg : 	Miles/(US) gallon mapped to x position, disp : displacement (cu.in.) to y position.
# Layer: points.

```

## B. Colour, Size, Shape and other aesthetic attributes

* ggplot have american and british spelling, both are works. ex : color and colour


### B.1 Aesthetics / aes () Attributes :

* aesthetics include some of the following:
    
  1. the x and y data arguments (mandatory)
    
  2. shapes
    
  3. colour
    
  4. fill
    
  5. size
    
  6. linetype
    
  7. alpha

* Different types of aesthetic attributes work better with different types of variables. 

* For example, colour and shape work well with categorical variables, while size works well for continuous variables. 

* The amount of data also makes a difference: if there is a lot of data it can be hard to distinguish different groups. An alternative solution is to use faceting, as described next.

#### B.1.1 x dan y data arguments

* x : data will be mapped on x axis

* y : data will be mapped on y axis

* x & y can be numerical or string data type depend on geom function will be used.

#### B.1.2 Shape



  
#### B.1.3 Colour

* colour set in aesthetic attributes would be scale based on their class.

* If you want to set an aesthetic to a fixed value, without scaling it, do so in the individual layer outside of aes().

```{r basic aesthetics attributes, include = False}

# basic graph with colour aes

ggplot(data, mapping = aes(x = mpg, y = disp, 
                           colour = cyl)) + geom_point()

# If you want to set an aesthetic to a fixed value, without scaling it, do so in the individual layer outside of aes(). Compare the following two plots:
ggplot(data, mapping = aes(x = mpg, y = disp, colour = "blue")) + geom_point()

ggplot(data, mapping = aes(x = mpg, y = disp)) + geom_point(colour = "blue")

# In the first plot, the value “blue” is scaled to a pinkish colour, and a legend is added. In the second plot, the points are given the R colour blue.

# more detail hit : https://ggplot2.tidyverse.org/articles/ggplot2-specs.html
```




## C. Faceting

* Faceting creates tables of graphics by splitting the data into subsets and displaying the same graph for each subset

* There are two types of faceting: grid and wrapped.

* To facet a plot you simply add a faceting specification with facet_wrap(), which takes the name of a variable preceded by ~. (ex : facewrap (~mpg))

```{r basic faceting, include=FALSE}
ggplot (data, mapping = aes(x = mpg, y = disp)) + geom_point(colour = 'red') + facet_wrap(~cyl)

```

## D. Geom Function

* geom: The name of the geometric object to use to draw each observation.Geoms can have additional arguments. All geoms take aesthetics as parameters. If you supply an aesthetic (e.g. colour) as a parameter, it will not be scaled, allowing you to control the appearance of the plot.
  
* list of argument that can be used in geom :
  
  1. size
  
  2. shape
  
  3. color / colour

* list of geom functions :

  1. geom_point() : fits a scatter plot (relationship plot) of x dan y values.
  
  2. geom_smooth() fits a smoother to the data and displays the smooth and its standard error.

  3. geom_boxplot() produces a box-and-whisker plot to summarise the distribution of a set of points.

  4. geom_histogram() and geom_freqpoly() show the distribution of continuous variables.

  5. geom_bar() shows the distribution of categorical variables.

  6. geom_path() and geom_line() draw lines between the data points. A line plot is constrained to produce lines that travel from left to right, while paths can go in any direction. Lines are typically used to explore how things change over time.


* stat: The name of the statistical tranformation to use. A statistical transformation performs some useful statistical summary, and is key to histograms and smoothers.

> You only need to set one of stat and geom: every geom has a default stat, and every stat a default geom.Most stats take additional parameters to specify the details of statistical transformation.

* position: The method used to adjust overlapping objects, like jittering, stacking or dodging.

* we will learn more detail at geom section.

## E. Modifying Axes

### E.1 Modifying labels
* you’ll learn the full range of options available in later chapters, but two families of useful helpers let you make the most common modifications. xlab() and ylab() modify the x- and y-axis labels:

```{r basic modifying axes, include=FALSE}
ggplot(data, mapping = aes(x = mpg, y = disp)) + geom_point() + 
  xlab ("this mpg") + ylab ("this disp")
```

### E.2 Rule or Limit Labels values.

* xlim() and ylim() modify the limits of axes:

```{r basic limit labels, include=FALSE}
ggplot(data, mapping = aes(x = mpg, y = disp)) + geom_point() + 
  xlab ("this mpg") + ylab ("this disp") +
  xlim(10,40) + ylim(80,400)

```



# 2. Layer

* It is useful to think about the purpose of each layer before it is added. In general, there are three purposes for a layer:

  1. To display the data. We plot the raw data for many reasons, relying on our skills at pattern detection to spot gross structure, local structure, and outliers. This layer appears on virtually every graphic. In the earliest stages of data exploration, it is often the only layer.

  2. To display a statistical summary of the data. As we develop and explore models of the data, it is useful to display model predictions in the context of the data. Showing the data helps us improve the model, and showing the model helps reveal subtleties of the data that we might otherwise miss. Summaries are usually drawn on top of the data.

  3. To add additional metadata: context, annotations, and references. A metadata layer displays background context, annotations that help to give meaning to the raw data, or fixed references that aid comparisons across panels. Metadata can be useful in the background and foreground.


## A. Basic Geom
 
 * Each of these geoms is two dimensional and requires both x and y aesthetics. 
 
 * All of them understand colour (or color) and size aesthetics,
 
 * some filled geoms (bar, tile and polygon) also understand fill. (not whole geom works with fill parameter)
 
  * geom_area() draws an area plot, which is a line plot filled to the y-axis (filled lines). Multiple groups will be stacked on top of each other.

  * geom_bar(stat = "identity") makes a bar chart. We need stat = "identity" because the default stat automatically counts value. The identity stat leaves the data unchanged. Multiple bars in the same location will be stacked on top of one another.

  * geom_line() makes a line plot. The group aesthetic determines which observations are connected. geom_line() connects points from left to right; geom_path() is similar but connects points in the order they appear in the data. Both geom_line() and geom_path() also understand the aesthetic linetype, which maps a categorical variable to solid, dotted and dashed lines. (geom path connected the data based on sequence of data)

* geom_point() produces a scatterplot. geom_point() also understands the shape aesthetic.

* geom_polygon() draws polygons, which are filled paths. Each vertex of the polygon requires a separate row in the data. It is often useful to merge a data frame of polygon coordinates with the data just prior to plotting.

* geom_rect(), geom_tile() and geom_raster() draw rectangles. geom_rect() is parameterised by the four corners of the rectangle, xmin, ymin, xmax and ymax. geom_tile() is exactly the same, but parameterised by the center of the rect and its size, x, y, width and height. geom_raster() is a fast special case of geom_tile() used when all the tiles are the same size.

* geom_text() adds text to a plot. It requires a label aesthetic that provides the text to display, and has a number of parameters (angle, family, fontface, hjust and vjust) that control the appearance of the text.

```{r basic geom, include=FALSE}
ggplot (data, mapping = aes(x = mpg, y = disp)) + geom_point() + geom_text(aes(label = mpg), family = "Times New Roman")

```

## B. Collective Geom

* An individual geom draws a distinct graphical object for each observation (row). For example, the point geom draws one point per row. 

* A collective geom displays multiple observations with one geometric object. This may be a result of a statistical summary, like a boxplot, or may be fundamental to the display of the geom, like a polygon.

* Lines and paths fall somewhere in between: each line is composed of a set of straight segments, but each segment represents two points. How do we control the assignment of observations to graphical elements? This is the job of the **group aesthetic.**

* by default, the **group aesthetic** is mapped to the interaction of all discrete variables in the plot.

* his often partitions the data correctly, but when it does not, or when no discrete variable is used in a plot, you’ll need to explicitly define the grouping structure by mapping group to a variable that has a different value for each group.

* If a group isn’t defined by a single variable, but instead by a combination of multiple variables, use **interaction()** to combine them, e.g. aes(group = interaction(school_id, student_id)).

### B.1 Multiple groups, one aesthetic

* in many situations, you want to separate your data into groups, but render them in the same way. In other words, you want to be able to distinguish individual subjects, but not identify them

* the following plot shows the growth trajectory for each boy (each Subject)

```{r collcetive geom, include=FALSE}
data(Oxboys, package = "nlme")

ggplot(Oxboys, mapping = aes(x = age, y = height)) + geom_point() + geom_line() # If you incorrectly specify the grouping variable, you’ll get a characteristic sawtooth appearance

ggplot(Oxboys, mapping = aes(x = age, y = height, group = Subject, color = Subject)) + geom_point() + geom_line()

# If a group isn’t defined by a single variable, but instead by a combination of multiple variables, use interaction() to combine them, e.g. aes(group = interaction(school_id, student_id)).
```

### B.2 Different groups on different layers

* Sometimes we want to plot summaries that use different levels of aggregation: one layer might display individuals, while another displays an overall summary. 

* tips : instead of setting the grouping aesthetic in ggplot(), where it will apply to all layers, we set it in geom_line() so it applies only to the lines

```{r different groups on different layers, include=FALSE}
# This is not what we wanted
ggplot(Oxboys, aes(age, height, group = Subject)) + 
  geom_line() + 
  geom_smooth(method = "lm", se = FALSE)

# instead of setting the grouping aesthetic in ggplot(), where it will apply to all layers, we set it in geom_line() so it applies only to the lines
ggplot(Oxboys, aes(age, height)) + 
  geom_line(aes(group = Subject)) + 
  geom_smooth(method = "lm", linewidth = 2, se = FALSE)
#> `geom_smooth()` using formula = 'y ~ x'
```

### B.3 Overriding the default grouping


