---
title: "ADVANCED R"
author: "M Adi Setiawan"
date: "2023-09-15"
output: html_document
---

> disclaimer : this article adopt from advanced r by Hadley Wickham



# 1. DATA STRCUTURE

There are 3 common data type that be used in data analys framework :

1. Homogenous Data : Atomic Vector, Matrix & n Array

2. Heterogenous Data : List, Data Frame

## 1.1 Vector

Three common properties in vector : 
  * typeof() : what it is
  
  * length() : how many elements it contains
  
  * attributes() : additional arbitrary metadata
  
  > all elements of an atomic vector must be the same type, whereas the elements of a list can have different types.
  
  > NB: is.vector() does not test if an object is a vector. Instead it returns TRUE only if the object is a vector with no attributes apart from names. Use is.atomic(x) || is.list(x) to test if an object is actually a vector.
  
  
### 1.1.1 ATOMIC VECTOR

Atomic vectors are usually created with c(), short for combine:

```{r atomic vector, include=FALSE}
dbl_var <- c(1, 2.5, 4.5)

# With the L suffix, you get an integer rather than a double 
int_var <- c(1L, 6L, 10L)

# Use TRUE and FALSE (or T and F) to create logical vectors 
log_var <- c(TRUE, FALSE, T, F)

chr_var <- c("these are", "some strings")

# Atomic vectors are always flat, even if you nest c()’s:
c(1, c(2, c(3, 4)))

# [1]1234
```


NA will always be coerced to the correct type if used inside c(), or you can create NAs of a specific type with : 

  * NA_real_ (a double vector) 
  * NA_integer_ 
  * NA_character_
  
(you can check through combining it with typeof() function)

### 1.1.1.1 Types and tests

Given a vector, you can determine its type with typeof(), or check if it’s a specific type with an “is” function: is.character(), is.double(), is.integer(), is.logical(), or, more generally, is.atomic().

> is.numeric() is a general test for the “numberliness” of a vector and returns TRUE for both integer and double vectors. It is not a specific test for double vectors, which are often called numeric.

```{r type test, include=FALSE}

typeof(int_var)
is.integer(int_var)
is.atomic(int_var)

typeof(dbl_var)
is.double(dbl_var)
is.atomic(dbl_var)

is.numeric(int_var)
is.numeric(dbl_var)

# NB : Numeric contain 2 type : integer : whole number (ex : 1,10) and double : possibly store decimal (ex : 1.11 , 2.00)

```

### 1.1.1.2 Coercion

All elements of an atomic vector must be the same type, so when you attempt to combine different types they will be coerced to the most flexible type.

```{r coercion vector, include=FALSE}

str(c("a", 1))
#>  chr [1:2] "a" "1" : stored as char

```

When a logical vector is coerced to an integer or double, TRUE becomes 1 and FALSE becomes 0. This is very useful in conjunction with sum() and mean().

```{r, include=FALSE}
x <- c(FALSE, FALSE, TRUE)
as.numeric(x)
str(as.numeric(x))

# so we can apply some basic math operating :

# Total number of TRUEs
sum(x) 
#>[1]1

# Proportion that are TRUE
mean(x)
#> [1] 0.3333

```

Coercion often happens automatically. Most mathematical functions (+, log, abs, etc.) will coerce to a double or integer, and most logical operations (&, |, any, etc) will coerce to a logical. You will usually get a warning message if the coercion might lose information. If confusion is likely, explicitly coerce with as.character(), as.double(), as.integer(), or as.logical().

### 1.1.2 List

Lists are different from atomic vectors because their elements can be of any type, including lists. You construct lists by using list() instead of c():

```{r list, include=FALSE}
x1 <- list(1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9))

str(x1)

```

Lists are sometimes called recursive vectors, because a list can con- tain other lists. This makes them fundamentally different from atomic vectors.

```{r recursive vector, include=FALSE}

x2 <- list(list(list(list()))) 

str(x2)

#>Listof1
#> $ :List of 1
#>   ..$ :List of 1
#>   .. ..$ : list()

is.recursive(x2)
#> [1] TRUE
```

c() will combine several lists into one. If given a combination of atomic vectors and lists, c() will coerce the vectors to list before combining them. Compare the results of list() and c():

```{r combination of atomic vector and list, include=FALSE}

x3 <- list(list(1, 2), c(3, 4)) 
str(x3)

y <- c(list(1, 2), c(3, 4)) 
str(y)

```

The typeof() a list is list. You can test for a list with is.list() and coerce to a list with as.list(). You can turn a list into an atomic vector with unlist(). If the elements of a list have different types, unlist() uses the same coercion rules as c().

## 1.2 Atrributes

* All objects can have arbitrary additional attributes, used to store meta data about the object.

* Attributes can be thought of as a named list (with unique names)

* Attributes can be accessed : 
  
  *individually with : attr()

  * all at once (as a list) with : attributes().

```{r attr and attributes, include=FALSE}
y2 <- 1:10

attr(y2, "my_attribute") <- "This is a vector" 

attr(y2, "my_attribute")
#> [1] "This is a vector"

str(attributes(y2))
#>Listof1
#> $ my_attribute: chr "This is a vector"

```

The only attributes not lost are the three most important:

  * Names, a character vector giving each element a name
  
  * Dimensions, used to turn vectors into matrices and arrays
  
  * Class,used to implement the S3objectsystem

Each of these attributes has a specific accessor function to get and set values. When working with these attributes, use names(x), class(x), and dim(x), not attr(x, "names"), attr(x, "class"), and attr(x, "dim").


### 1.2.1 Names

* You can name a vector in three ways: (function : names () and setnames () )

  1.  When creating it: x<-c(a=1,b=2,c=3).

  2. By modifying an existing vector in place: x <- 1:3; names(x) <- c("a", "b", "c").
  3. By creating a modified copy of a vector: x <- setNames(1:3, c("a", "b", "c")).

Not all elements of a vector need to have a name. If some names are missing, names() will return an empty string for those elements. If all names are missing, names() will return NULL.

```{r names, include=FALSE}

y3 <- c(a=1,2,3)
names(y3)

# or this setNames(y3, c('a'))

#> [1] "a" "" ""

z<-c(1,2,3)
names(z)
#> NULL

```

### 1.2.2 Factors

* One important use of attributes is to define factors. 

* A factor is a vector that can contain only predefined values, and is used to store categorical data. 

* Factors are built on top of integer vectors using two attributes :
  
  1. the class(), “factor”, which makes them behave differently from regular integer vectors.
  
  2. the levels(), which defines the set of allowed values.

> common function used on factor : factor (), class (), levels()

```{r factor, include=FALSE}

x1 <- factor(c("a", "b", "b", "a"))
x1
#>[1]abba
#> Levels: a b
class (x1)
#> [1] "factor"
levels(x1)
#> [1] "a" "b"

# You can't use values that are not in the levels
x1[2] <- "c"
#> Warning: invalid factor level, NA generated
x1
#>[1]a <NA>b a
#> Levels: a b

```

Factors are useful when you know the possible values a variable may take, even if you don’t see all values in a given dataset. Using a factor instead of a character vector makes it obvious when some groups contain no observations:

```{r factor case, include=FALSE}
sex_char <- c("m", "m", "m")
sex_factor <- factor(sex_char, levels = c("m", "f"))
table(sex_char) 
#> sex_char
#>m
#>3 table(sex_factor) 
#> sex_factor 
#>m f
#>3 0
```

* Sometimes when a data frame is read directly from a file, a column you’d thought would produce a numeric vector instead produces a factor. 

* This is caused by a non-numeric value in the column, often a missing value encoded in a special way like . or -. 

* To remedy the situation :

  1. coerce the vector from a factor to a character vector
  
  2. and then from a character to a double vector. 
  
* Be sure to check for missing values after this process.
  
* of course, a much better plan is to discover what caused the problem in the first place and fix that ; using the na.strings argument to read.csv() is often a good place to start.

```{r fixing factor while read data from csv, include=FALSE}

z1 <- read.csv(text = "value\n12\n1\n.\n9")

typeof(z1$value)
#> [1] "integer"

as.double(z1$value)
#>[1]3214
# Oops, that's not right: 3 2 1 4 are the levels of a factor, # not the values we read in!

class(z1$value)
#> [1] "factor"

#Wecanfixitnow:
as.double(as.character(z1$value))

#> Warning: NAs introduced by coercion
#>[1]12 1NA 9

# Or change how we read it in:
z1 <- read.csv(text = "value\n12\n1\n.\n9", na.strings=".") 
typeof(z1$value)
#> [1] "integer"

class(z1$value)
#> [1] "integer" z$value
#>[1]12 1NA 9 # Perfect! :)
```

While factors look (and often behave) like character vectors, they are actually integers. Be careful when treating them like strings. Some string methods (like gsub() and grepl()) will coerce factors to strings, while others (like nchar()) will throw an error, and still others (like c()) will use the underlying integer values. For this reason, it’s usually best to explicitly convert factors to character vectors if you need string-like behaviour. In early versions of R, there was a memory advantage to using factors instead of character vectors, but this is no longer the case.

## 1.3 Matrices and Array

* Common function that used to be on matrix and array 

|                         |  matrix                 | array             |
|:------------------------|:-----------------------:|:-----------------:|
|structure                |two dimensions           |Multidimensional   |
|Assignment form          |matrix(),dim()           |array(),dim()      |
|length (ndata)           |length()                 |length()           |
|nrow                     |nrow(),ncol()            |dim()              |
|assigning row & col name |rownames(),colnames()    |dimnames()         |
|binding data             |rbind(),cbind()          |abind()            |
|check matrix or array    |is.matrix()              |is.array()         |
|assign as                |as.matrix()              |as.array()         |
|matrix transpose         |t()                      |t()                |



```{r matrix and array, include=FALSE}

# create matrix and array
matrix_a <-  matrix(data = seq (1,6,1), nrow = 2, ncol = 3)
array_a <-  array(1:6, dim = c(2,3,2)  ) # dim = c (row,col,list no)

# or we can use dim to convert existing object to matrix and array
a <- 1:6
# creating matrix with dim()
dim(a) <- c(2,3)
a

# creating array with dim ()
dim(a) <- c(2,3,1)
a

# view matrix and array structure
dim(matrix_a) # or
dim (array_a)

# view length of whole data (ndata)
length (matrix_a)
length (array_a)

# view length of row data or col (nrow or ncol)

# on matrix and array
nrow(matrix_a)
ncol (array_a)


# assign names of row and col (for indexing row and col) of matrix and array
# for matrix, use rownames() and colnames()
rownames(matrix_a) <- c("a","b")
colnames(matrix_a) <- c("c","d", "e")
matrix_a

# for array, use dimnames()

dimnames(array_a) <- list(c("a","b"), c("c",'d','e'), c('A','B'))
array_a

array_a["A"]

```


```{r binding then check then assign as, include=FALSE}
add_b <- seq (7,12,1)
dim (add_b) <- c(2,3)

rbind(matrix_a, add_b)

# check is it matrix ?
is.matrix(array_a)

# check is it array ?
is.array(matrix_a)

# assign as matrix

as.matrix(add_b)
is.matrix(add_b)


# transpose matrix

t(add_b)

```

> Vectors are not the only 1-dimensional data structure. You can have matrices with a single row or single column, or arrays with a single. see the differences with str () function.


## 1.4 data Frame

data frame is a list of equal-length vectors. This makes it a 2-dimensional structure, so it shares properties of both the matrix and the list.

|                         |  Dataframe              | 
|:------------------------|:-----------------------:|
|structure                |two dimensions           |
|Assignment form          |data.frame()             |
|length (ndata)           |length()                 |
|nrow                     |nrow(),ncol()            |
|assigning row & col name |rownames(),colnames()    |
|binding data             |rbind(),cbind()          |
|check matrix or array    |is.data.frame()          |
|assign as                |as.data.frame()          |
|transpose                |t()                      |




```{r dataframe, include=FALSE}

# create a dataframe
x <- data.frame(x = seq(1,3,1), y = c('a','b','c'))
str(x)

```


### 1.4.1 Testing & Coercion

* Because a data.frame is an S3 class, its type reflects the underlying vector used to build it : the list. 

* To check if an object is a data frame, use class() or test explicitly with is.data.frame()

* to coerce a data frame use as.data.frame ()
  
  * A vector will create a one-column data frame.
  
  * A list will create one column for each element; it’s an error if they’re
not all the same length.
  
  * A matrix will create a data frame with the same number of columns
and rows.


```{r test & coercion data.frame, include=FALSE}
# TESTING

typeof(x)
#> "list"

class (x)
#> "data.frame"

is.data.frame(x)
#> TRUE


## COERCE : you can coerce an object to data frame with : as.data.frame ()

y <- c("x" = c(1,2,3),
       "y" = c('a','b','c'))
class(y)

#> "character"

y1 <- as.data.frame(y)
class(y1)
#> "data.frame"

```


### 1.4.2 Combining Data Frames

* we can combine a data frame using cbind() and row cbind() :

```{r cbind, include=FALSE}
cbind(x, z=c(3:1))

rbind(x, z = c(3:2))


```

Key take away of bind row and column of dataframe :

* When combining column-wise, the number of rows must match, but row names are ignored.

* When combining row-wise, both the number and names of columns must match

* Use "plyr::rbind.fill()" to combine data frames that don’t have the same columns.

* avoid bind a factor data type.

> The conversion rules for cbind() are complicated and best avoided by ensuring all inputs are of the same type.

### 1.4.3 Special Columns

* Since a data frame is a list of vectors, it is possible for a data frame to have a column that is a list

```{r special column, include=FALSE}

df <- data.frame(x = 1:3)
y <- list(1:2, 1:3, 1:4)

# list add as whole in single cell ()
df$y <- y
df [2]

# it tries to put each item of the list into its own column

# data.frame(x = 1:3, y = list(1:2, 1:3, 1:4))
#> arguments imply differing number of rows: 2, 3, 4

# A workaround is to use I(), which causes data.frame() to treat the list as one unit:

# I () stand for : Change the class of an object to indicate that it should be treated ‘as is’.
df1 <- data.frame(x = 1:3, y = I(list(1:2, 1:3, 1:4)))
str(df1)

```

* Similarly, it’s also possible to have a column of a data frame that’s a matrix or array, as long as the number of rows matches the data frame:

```{r dataframe combined matrix, include=FALSE}
df2 <- data.frame(x = 1:3, 
                  y = I(matrix(data = seq(1,6,1), ncol = 2, nrow = 3)))

str(df2)
df2[2,'y']

```

## 2 SUBSETTING

* The three subsetting operators.

* The six types of subsetting.

* Important differences in behaviour for different objects (e.g., vectors, lists, factors, matrices, and data frames).

* The use of subsetting in conjunction with assignment.

* subsetting operators : [], [[]], $



### 2.1 Subsetting atomic vector

* There are five things that you can use to subset a vector :
  
  1. positive integers : return elements at the specified positions
  
  2. negative integers : omit elements at the specified positions (You can’t mix positive and negative integers in a single subset)
  
  3. Logical: select elements where the corresponding logical value is TRUE
  
  4. NULL ([]) : returns the original vector
  
  5. Zero : Zero returns a zero-length vector. usually used in test data.
  
  6. Character : return elements with matching names

```{r subsetting atomic vector, include=FALSE}
x1 <- seq (1,10,1)

# 1. positive integers : return element in spesific positions
x1 [4:8]
#> [1] 4 5 6 7 8

x1 [c(10,2,6)]
#> [1] 10  2  6

# Duplicated indices yield duplicated values
x1[c(2, 2)]
#> [1] 2 2

# Real numbers are silently truncated to integers
x1[c(2.1, 2.9)]
#> [1] 2 2

# 2. negative integers : omit/remove element in spesific positions

x1 [-1]
#> [1]  2  3  4  5  6  7  8  9 10


x1 [c(-2,-10)]
#> [1]  1  3  4  5  6  7  8  9 

# You can’t mix positive and negative integers in a single subset

# x1[c(-1, 2)]
#> Error in x1[c(-1, 2)] : only 0's may be mixed with negative subscripts

# 3. Logical Vector : return element on TRUE logical positions

x1 [c(T,F,T,F)] # TRUE stand for odd number (angka ganjil)
#> [1] 1 3 5 7 9

# If the logical vector is shorter than the vector being subsetted, it will be recycled to be the same length.

# above equivalent as below :
x1 [c(T,F)]
#> [1] 1 3 5 7 9

#> [1] 1 3 5 7 9
x1 [x1>3]
#> [1]  4  5  6  7  8  9 10

x1 [x1%%2 == 0] # return whole number (angka genap)
#> [1]  2  4  6  8 10

# A missing value in the index always yields a missing value in the output:
x1 [c(T,F,NA,T)]
#> [1]  1 NA  4  5 NA  8  9

# 4. Nothing : return the orignal vector

x1 []
#> 1  2  3  4  5  6  7  8  9 10

# 5. Zero : Zero returns a zero-length vector
x1[0]
#> numeric(0)

x2 <- c('a','b','c')
x2 [0]
#> character(0)

# 6. Character Vector : return elements with matching names

y<- setNames(object = x1, letters[1:10])

y[c('a','j','d')]
#>  a  j  d 
#>  1 10  4 

# Like integer indices, you can repeat indices
y[c('e','e','e')]
#>  e  e  e 
#>  5  5  5  

# When subsetting with [ names are always matched exactly. if any match it will return NA

z <- c('abc' = 1, 'def'=2)
z [c('a','d')]
#> <NA> <NA> 
#>  NA   NA

z[c('abc','def')]
#> abc def 
#> 1   2 



```








